---
title: "jacob_martin_w03_assignment"
output: html_document
date: "2024-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning= FALSE, echo = TRUE)
```

## Scatter Plot - An Overview


The following graph shows the differences in flipper length vs bill length, based off the species of a given penguin.

```{r penguin_data_no_overlay, echo=TRUE}
# External Libs.
library(ggplot2)
library(jpeg)

# Read in required metadata.
data <- read.csv('W03-Penguins.csv')
img <- readJPEG('penguin.jpeg')

# Plot 
gg <- ggplot(data, aes(x = bill_length_mm, y = flipper_length_mm, color = species, size = body_mass_g)) +
  # As Scatter
  geom_point() +
  # Label Accordingly,
  labs(x = "Bill Length (MM)", y ="Flipper Length (MM)", title="Penguin Species with Flipper Length vs. Bill Length and Body Size") +
  # Set the colors based off species
  scale_color_manual(values = c("Adelie" = "blue", "Gentoo"="green","Chinstrap" = "red")) + 
  # Scale literal data point based off body mass
  scale_size_continuous(range = c(0, 3)) +
  guides(color = guide_legend(title = "Species"), size = guide_legend(title = "Body Mass (g)"))

print(gg)
```

Notice how the Adelie species has consistently the smallest bills and flipper size. While the Chinstrap species lands as the the second largest bills and flippers, and lastly the Gentoo has consistently the largest bills and flippers. It's also worth noting that the Gentoo have the largest body mass indicated by the size of the data point on the graph. 

## Scatter Plot - Img Overlay

Below is a scatterplot which has an image overlay on it. 

```{r penguin_data, echo=FALSE}
# External Libs.
library(ggplot2)
library(jpeg)

# Read in required metadata.
data <- read.csv('W03-Penguins.csv')
img <- readJPEG('penguin.jpeg')

# Plot 
gg <- ggplot(data, aes(x = bill_length_mm, y = flipper_length_mm, color = species, size = body_mass_g)) +
  # Add a background image
  ggpubr::background_image(img) +
  # As Scatter
  geom_point() +
  # Label Accordingly,
  labs(x = "Bill Length (MM)", y ="Flipper Length (MM)", title="Penguin Species with Flipper Length vs. Bill Length and Body Size") +
  # Set the colors based off species
  scale_color_manual(values = c("Adelie" = "blue", "Gentoo"="green","Chinstrap" = "red")) + 
  # Scale literal data point based off body mass
  scale_size_continuous(range = c(0, 3)) +
  guides(color = guide_legend(title = "Species"), size = guide_legend(title = "Body Mass (g)"))

print(gg)
```

While the values of the data between the two grahps are the same, this one depicts how we can utilize an image overlay when graphing some data.
